Pretty darn good control: when are approximate solutions better than approximate models

### 2023-01-19 manuscript/manuscript.Rmd
:pen:

In this section we introduce some models describing the population dynamics of fisheries.
In general, the class of models that appear in this context are *first order finite difference equations*.
For $n$ species, these models have the general form
\begin{align}
  \label{eq:general model}
  N_{t+1} - N_t = f(N_t) - h_t := \Delta N_t,
\end{align}
where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, and $h_t$ is the harvest size at timestep $t$.

@@

Most commonly, fisheries use 1-dimensional models, $n=1$.
One such model appearing in a variety of ecological contexts is *logistic growth*, for which
$$
  f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
$$
More relevant to our current work, is the following single-species model studied in \cite{may77},
\begin{align}
  \label{eq:may}
  \Delta X_t
  = L(X_t;\ r, K) - F(X_t, H; \beta,c) - h_t,
\end{align}
where,
\begin{align*}
  F(X,H;\ \beta,c) := \frac{\beta H X^2}{c^2 + X^2}.
\end{align*}
The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal rate of predation $\beta$, and the predation half-maximum biomass $c$.

Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (see Fig. \ref{tbd}).
Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $V_1$ of the system, these two attractors separated by an unstable fixed point.
At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
For this value of $H$, and for higher values, only the lower fixed point remains.

This structure implies two things.
First, that a drift in $H$ could lead to catastrophic consequences, with the population $V_1$ plummeting to the lower fixed stable point.
Second, that if the evolution of $V_1$ is *stochastic*, then, even at values of $H$ below the threshold point, the system runs a sizeable danger of tipping over towards the lower stable point.

@@

Consider the following three species *generalization* of eq. \eqref{eq:may}:
\begin{align}
\label{eq:3d model}
\begin{split}
  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t,\\
  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
\end{split}
\end{align}
The model is conceptualized in Fig. \ref{fig:tbd}.
It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
There are ten parameters in this model: 
The growth rate and carrying capacity, $r_X$, $K_X$, $r_Y$ and $K_Y$, of $X$ and $Y$.
A parameter $c_{XY}$ mediating a Lotka-Volterra competition between $X$ and $Y$.
A maximum hunting intensity $\beta$ and half-maximum $c$ specifying how $Z$ forages on $X$ and $Y$.
A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
Finally, a death rate $d_Z$ and a parameter $f$ related to the birth rate of $Z$.

@@

The title of this piece references a mathematical biology workshop at NIMBioS
organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
2011 seeking to pivot control from 'optimal' to 'pretty darn good'.

>>>>

In this section we introduce some models describing the population dynamics of fisheries.
In general, the class of models that appear in this context are *first order finite difference equations* (potentially stochastic difference equations).
For $n$ species, these models have the general form

\begin{align}
  \label{eq:general model}
  N_{t+1} - N_t = f(N_t) - F N_t := \Delta N_t,
\end{align}
where $N_t = (X_t,\ Y_t,\ \dots) \in \mathbb{R}^{n}_+$ is a vector of populations, $f:\mathbb{R}^{n}\to\mathbb{R}^{n}$ is an arbitrary function, where the product of the mortality and the stock size, $F N_t$ is the harvest size at timestep $t$.

@@

Optimal control policies for fisheries are frequently based on 1-dimensional models, $n=1$.
The most familiar model of $f(X)$ is that of *logistic growth*, for which

$$
  f(X_t) = r X_t\big(1 - X_t / K \big) =: L(X_t;\ r, K).
$$
Which combined with the harvest term from \eqref(eq:general model) is known as the Gordon-Schaefer model, after foundational work which first developed the concept of maximum sustainable yield, MSY, [@Gordon1954; @Schaefer1954], which remains a central concept in fisheries management.  
In this model, note that the maximum growth rate occurs at population size of $N = K/2$, which is approached asymptotically from any positive initial state $N_0 \in (0,K)$ under a constant mortality of $F = r/2$, which thus produces a harvest of $rK/4$ at equilibrium -- the maximum yield sustainable indefinitely.

Real world ecological systems are obviously far more complicated than this simple model suggests.  
One particularly important aspect that has garnered much attention is the potential for the kind of highly non-linear functions that can support dynamics such as alternative stable states and hysteresis.
A seminal example of such dynamics was introduced in \cite{may77}, using a one-dimensional model of a prey (resource) species under the pressure of a (fixed) predator:

\begin{align}
  \label{eq:may}
  f(X_t)
  = L(X_t;\ r, K) - F(X_t, H; \beta,c),
\end{align}
where,
\begin{align*}
  F(X_t,H;\ \beta,c) := \frac{\beta H X_t^2}{c^2 + X_t^2}.
\end{align*}

The model has five parameters: the growth rate $r$ and carrying capacity $K$ for $X$, a constant population $H$ of a species which preys on $X$, the maximal rate of predation $\beta$, and the predation half-maximum biomass $c$.

Eq. \eqref{eq:may} is an interesting study case of a *tipping point* (saddle-node bifurcation) (see Fig. \ref{tbd}).
Holding the value of $\beta$ fixed, for intermediate values of $H$ there exist two stable fixed points for the state $\hat X_t$ of the system, these two attractors separated by an unstable fixed point.
At a certain threshold value of $H$, however, the top stable fixed point collides with the unstable fixed point and both are annihilated.
For this value of $H$, and for higher values, only the lower fixed point remains.
This also creates the phenomenon of *hysteresis*, where returning $H$ to its original value is not sufficient to restore $\hat X_t$ to the original stable state.

This structure implies two things.
First, that a drift in $H$ could lead to catastrophic consequences, with the population $X_t$ plummeting to the lower fixed stable point.
Second, that if the evolution of $X_t$ is *stochastic*, then, even at values of $H$ below the threshold point, the system runs a sizeable danger of tipping over towards the lower stable point.

<!-- Discuss control solutions of this model(?) -->
<!-- Briefly motivate reason to consider 3d model, adding competition and predation processes -->

@@

Consider the following three species *generalization* of eq. \eqref{eq:may}:

\begin{align}
\label{eq:3d model}
\begin{split}
  \Delta X_t &= L(X_t;\ r_X, K_X) - F(X_t, Z_t;\ \beta, c) - c_{XY}X_tY_t - h_t,\\
  \Delta Y_t &= L(Y_t;\ r_Y, K_Y) - D F(Y_t, Z_t;\ \beta, c) - c_{XY}X_tY_t,\\
  \Delta Z_t &= \left(f(X_t + DY_t\right) - d_Z)Z_t.
\end{split}
\end{align}
The model is conceptualized in Fig. \ref{fig:tbd}.
It contains three populations, the state at time $t$ is $(X_t, Y_t, Z_t)$. 
Species $Z$ preys on both $X$ and $Y$, while the latter two compete for resources.
There are ten parameters in this model: 
The growth rate and carrying capacity, $r_X$, $K_X$, $r_Y$ and $K_Y$, of $X$ and $Y$.
A parameter $c_{XY}$ mediating a Lotka-Volterra competition between $X$ and $Y$.
A maximum hunting intensity $\beta$ and half-maximum $c$ specifying how $Z$ forages on $X$ and $Y$.
A parameter $D$ regulating a relative preference of $Z$ to prey on $Y$.
Finally, a death rate $d_Z$ and a parameter $f$ related to the birth rate of $Z$.

@@

The title of this piece references a mathematical biology workshop at NIMBioS
organized by Paul Armsworth, Alan Hastings, Megan Donahue, and Carl Towes in
2011 which first sought to emphasize 'pretty darn good' control solutions to
more realistic problems over optimal control to idealized ones. 
This material is based upon work supported by the National Science Foundation under Grant No. DBI-1942280.
